\documentclass[12pt]{article}
\usepackage{doi}
\PassOptionsToPackage{hyphens}{url}
\usepackage{amsfonts,amsmath,extarrows,amssymb}
\title{Cantor-Bernstein for Theories}
\author{Princeton Logic Group}
\usepackage{amsthm}
\usepackage[natbib=true,style=chicago-authordate,sorting=nyt]{biblatex}
% \usepackage{biblatex-chicago}
\addbibresource{/Users/hhalvors/org/ref/master.bib}
\newtheorem{prop}{Proposition}
\newtheorem{fact}[prop]{Fact}
\newtheorem{cor}[prop]{Corollary}
\newtheorem{conj}[prop]{Conjecture}
\newtheorem{lemma}[prop]{Lemma}
\theoremstyle{definition}
\newtheorem*{defn}{Definition}
\newtheorem{example}[prop]{Example}
\theoremstyle{remark}
\newtheorem*{note}{Remark}
\newtheorem*{question}{Question}

\newcommand{\3}{\mathcal}
\begin{document}

%% TO DO: Oligomorphic groups

%% TO DO: Ahlbrandt and Ziegler results

\maketitle

The purpose of this note is to ask: under what conditions could a pair
of theories $(T_1,T_2)$ fail to have the Cantor-Bernstein or
co-Cantor-Bernstein property?

\begin{defn} We say that the pair $(T_1,T_2)$ has the Cantor-Bernstein
  property just in case: if $T_1$ and $T_2$ are mutually faithfully
  interpretable, then $T_1$ and $T_2$ are bi-interpretable. In other
  words: if $(T_1,T_2)$ does not have the Cantor-Bernstein property
  then (a) $T_1$ and $T_2$ are mutually faithfully interpretable, and
  (b) $T_1$ and $T_2$ are not bi-interpretable. \end{defn}

Here ``mutually faithfully interpretable'' means that there are
conservative (strong, equality preserving) translations $F:T_1\to T_2$
and $G:T_2\to T_1$. It would also be interesting to look at this
question for the case of more general translations, with and without
equality preservation. Could we end up getting different kinds of
answers in the two cases? Or is the latter case reducible, in some
sense, to the former?

\begin{defn} We say that the pair $(T_1,T_2)$ has the
  co-Cantor-Bernstein property just in case: if there are essentially
  surjective translations $F:T_1\to T_2$ and $G:T_2\to T_1$, then
  $T_1$ and $T_2$ are bi-interpretable. \end{defn}

\newcommand{\2}{\mathsf}

\section{Factorization of translations}

For strong translations, we have the following factorization theorem:
\begin{quote} A translation $F$ is an equivalence iff $F$ is
  conservative and essentially surjective \citep[Props.\ 4.5.26,
  4.5.27]{tlps}. \end{quote} This fact is useful for reasoning about
relations between theories: if $T_1$ can be embedded into $T_2$ such
that every $\Sigma _2$-formula is equivalent to a formula in the
image, then $T_1$ and $T_2$ are equivalent. It suggests the validity
of the following mode of reasoning about the relationship between
theories:
\begin{quote}
  If $T_1$ can be interpreted into $T_2$ into such a way that (a) any
  inference licensed by $T_2$ was already licensed by $T_1$, and (b)
  any concept expressable by $T_2$ is already expressable by $T_1$,
  then $T_1$ and $T_2$ are equivalent. \end{quote} The purpose of this
section is to investigate the extent to which this kind of reasoning
works for translations in general (and not just strong
translations). In fact, there is a wide-open field of investigation
here, which should hold much interest for talk about ``relations
between theories.'' In particular, with the generalized notion of
translations, first-order theories form not just a category but a
$2$-category. In a category $C$, the interesting properties of
morphisms include: iso, monic, epi, regular monic, split monic,
etc. Furthermore, the interesting technical questions include: if a
morphism $f:a\to b$ is both epi and monic, then is it is?  Hence, if
theories formed only a $1$-category, then we might try to cache out
notions like ``equivalence'' and ``reduction'' in terms of notions
like ``isomorphism'' and ``monomorphism.''

But theories form a $2$-category, and so different features of
morphisms become relevant. Think, for example, of the paradigm case of
a $2$-category, viz.\ the category $\2{Cat}$ of all (small)
categories. Here the interesting properties of morphisms (functors)
are not ``is a monomorphism'' and ``is an epimorphism'' etc., but ``is
full'' and ``is faithful'' and ``is essentially surjective'' etc. (For
a general $2$-category $D$, the corresponding properties of morphisms
have slightly more complicated definitions.)  Similarly, the
interesting notion of ``sameness of objects'' in $\2{Cat}$ is not
isomorphism, but equivalence of categories --- which can be thought of
as ``isomorphism up to isomorphism.'' Finally, for the $2$-category of
theories, the interesting question is not whether $f$ is an
isomorphism iff $f$ is a monomorphism and epimorphism, but whether $f$
is an equivalencee iff $f$ is [some other kind of thing]. See
\url{https://ncatlab.org/nlab/show/%28eso%2C+fully+faithful%29+factorization+system}
  for more on factorization of morphisms in $2$-categories.                                

In short: if relations between theories are to be explicated in terms
of properties of translations, then these explications should make use
of $2$-categorical concepts.

Note that there are actually two distinct $2$-categories at play here:
the ``syntactic'' $2$-category $\2{Th}$ of theories and translations,
and the ``semantic'' $2$-category $\2{Mod}$ of categories of models of
theories. The interesting properties of morphisms will be different in
the two cases, but of course, also related to --- in fact, dual to ---
each other, since there is a contravariant $2$-functor
$M:\2{Th}\to\2{Cat}$ that takes a theory $T$ to its category $M(T)$ of
models, a translation $F:T_1\to T_2$ to the dual functor
$M(F):M(T_2)\to M(T_1)$, and a t-map $\chi : F\Rightarrow G$ to a
natural transformation $M(\chi ):M(F)\Rightarrow M(G)$.

\begin{defn} Let $F:T_1\to T_2$ be a translation. We say that
  $F:T_1\to T_2$ is \emph{conservative} just in case if
  $T_2\vdash F\phi$ then $T_1\vdash \phi$, for any
  $\Sigma _1$-sentence $\phi$. \end{defn}

\begin{question} Is the above a good definition for coherent theories?
  Recall that the implications of a coherent theory cannot be captured
  by sentences, but need (open) sequents. So should the definition of
  ``conservative'' involve sequents?

  In terms of syntactic categories: the above definition only involves
  subobjects of $1$. However, the notion of conservativity should
  involve all subobject lattices. \end{question}

The following example shows that in the $2$-category of theories, a
translation's being conservative and essentially surjective does not
imply that it is an equivalence.

\begin{example} \label{nope} We describe a translation
  $F:(T\cup T')\to T$ that is conservative and essentially surjective,
  but not an equivalence. Let $\Sigma$ be a signature with a single
  sort $\sigma$, and let $T$ be an arbitrary complete theory in
  $\Sigma$. (For example, we could take $T$ to be the theory that says
  that there are infinitely many things. Or we could take $T$ to be
  the theory that says that there are exactly two things.)  Let $T'$
  be a primed copy of $T$, and let $T\cup T'$ be the coproduct of $T$
  and $T'$. That is, the signature of $T\cup T'$ is the disjoint union
  $\Sigma \cup \Sigma '$, and the axioms of $T\cup T'$ consist of the
  original axioms of $T$ along with a primed copy of those
  axioms. Note that any $\Sigma\cup\Sigma '$-formula is provably
  equivalent to either a conjunction or a disjunction of two formulas
  $\phi ,\phi '$, the first a $\Sigma$-formula and the second a
  $\Sigma '$-formula. What's more, if
  $(T\cup T')\vdash \phi \vee \phi '$ then either
  $(T\cup T')\vdash\phi$ or $(T\cup T')\vdash \phi '$, since any pair
  of models of $T$ canonically defines a model of $T\cup T'$.

  Now define an equality-preserving reconstrual $F$ from
  $\Sigma\cup \Sigma '$ to $\Sigma$ by removing the primes from
  symbols in $\Sigma '$. We observe:
  \begin{itemize}
  \item $F$ is a translation from $T\cup T'$ to $T$. In the case of a
    $\Sigma$-formula $\phi$, it is obvious that if
    $T\cup T'\vdash\phi$ then $T\vdash\phi$. For the more general
    case, any $\Sigma\cup \Sigma '$-formula $\psi$ is provably
    equivalent either to $\phi \vee \phi '$ or to $\phi \wedge\phi '$,
    where $\phi$ is a $\Sigma$-formula and $\phi '$ is a
    $\Sigma '$-formula. In the former case, if
    $T\cup T'\vdash \phi \vee \phi '$, then either
    $T\cup T'\vdash \phi$ or $T\cup T'\vdash \phi '$, and hence either
    $T\vdash F\phi$ or $T\vdash F\phi '$.
  \item $F$ is essentially surjective.
  \item $F$ is conservative. This follows immediately from the fact
    that $T\cup T'$ is complete.
  \item $F$ is not part of an equivalence. If $F$ had a
    $1$-dimensional inverse $G:T\to (T\cup T')$, then $GF$ would map
    both sorts $\sigma$ and $\sigma '$ to the same sort. Suppose for
    simplicity that $GF(\sigma )=GF(\sigma ')=\sigma$. If there were a
    t-map $\chi :GF\Rightarrow 1$, then $\chi _{\sigma '}$ would be a
    definable bijection of arity $\sigma\to\sigma '$, in contradiction
    with the fact that $\Sigma \cup \Sigma '$ has no such functional
    relations. (TO BE COMPLETED. I have the strange feeling that $F$
    could be an equivalence if $T$ has only infinite models, and if
    $G$ is $2$-dimensional with $G(\sigma )=\sigma ,\sigma '$. For
    example, what if $T$ is the theory that says that there are
    infinitely many things? In that case, aren't two copies of $T$
    equivalent to one copy? But surely if $T$ is the theory that says
    ``there are exactly two things'', then $GF$ would not be homotopy
    equivalent to the identity.)
  \end{itemize}
  The reason why $F$ is not an equivalence is because neither
  conservativity nor essential surjectivity require that the domain
  theory has a sufficient number of functional relations between
  sorts. In this case, $T$ describes a functional relation (namely the
  identity) between sorts $F(\sigma )$ and $F(\sigma )$, whereas
  $T\cup T'$ does not describe a functional relation between $\sigma$
  and $\sigma '$.  The issue then seems to be that $F$ does not
  resemble a full functor. Intuitively, a full functor would have the
  feature that for any functional relation
  $\chi :F(\sigma )\to F(\sigma ')$, there is a functional relation
  $\theta :\sigma \to \sigma '$ such that $F(\theta )\simeq
  \chi$. While there is indeed a $\Sigma\cup \Sigma '$-formula $\eta$
  such that $F(\eta )$ is provably equivalent to
  $\chi (x,y)\equiv (x=_\sigma y)$, any such $\eta$ has the wrong
  arity.

  In terms of the syntactic categories $C_{T\cup T'}$ and $C_{T}$,
  we can see that the functor corresponding to $F$ is not full. In
  particular, if $x$ is a variable of sort $\sigma$ and $x'$ is a
  variable of sort $\sigma '$, then there is an arrow from
  $[\top .Fx]$ to $[\top .Fx']$ in $C_{T}$, but no arrow from
  $[\top .x]$ to $[\top .x']$ in $C_{T\cup T'}$.  \end{example}

\begin{note} Here's a little puzzle: $T\cup T'$ is Morita equivalent
  to a single-sorted theory $T_1$.  Since $T\cup T'$ is proper (or may
  be assumed so), there is a translation $G:T_1\to T\cup T'$ that is
  one-half of an equivalence. Why is the composite $F\circ G$ not an
  equivalence? Isn't it a strong translation that is conservative and
  essentially surjective? \end{note}

\begin{example} Let
  $\Sigma _1 = \{ \sigma _1,\sigma _2,\chi _1,\chi _2\}$, where
  $\chi _i:\sigma _1\to \sigma _2$. Let $T_1$ be the theory in
  $\Sigma _1$ that says that $\chi _1$ and $\chi _2$ are functional
  relationships. Let $\Sigma _2 =\{ \sigma _1',\sigma _2',\theta \}$,
  and let $T_2$ be the theory in $\Sigma _2$ that says that $\theta$
  is a functional relationship. Let $F:\Sigma _1\to \Sigma _2$ be the
  reconstrual that takes both $\chi _1$ and $\chi _2$ to
  $\theta$. Then $F:T_1\to T_2$ is a translation, and as in the
  previous example, $F$ is essentially surjective. However, $F$ is not
  conservative. Indeed, we have
  $T_2\vdash F[\forall x\forall x(\chi _1\leftrightarrow \chi _2)]$
  but
  $T_1\not\vdash \forall x\forall y(\chi _1\leftrightarrow \chi _2)$.
\end{example}

Now here is my intuition: for a (weak) translation $F:T_1\to T_2$, the
notion of being conservative still makes sense --- although it may not
be a very powerful or useful notion. Indeed, to say that $F$ is
conservative basically says that if $F\phi (x)\vdash F\psi (x)$ then
$\phi (x)\vdash\psi (x)$, and that only tells us something about the
subobject lattices of the syntactic categories $C_{T_1}$ and
$C_{T_2}$. In contrast, to say that a logical functor from $C_{T_1}$
to $C_{T_2}$ is fully faithful means that for any fixed objects
$[\phi .X]$ and $[\psi .Y]$ of $C_{T_1}$, the map
\[ \mathrm{hom}\left( [\phi .X] ,[\psi .Y]\right) \: \to \:
  \mathrm{hom}\left( [F\phi .X] ,[F\psi .Y]\right) ,\] is a
bijection. The fullness condition implies that the translation is
conservative: if $F(\phi )\vdash F(\psi )$ in $T_2$, then $[X=X]$ is
an arrow from $F(\phi )$ to $F(\psi )$ in $C_{T_2}$. By the fullness
of $F$, there is a corresponding arrow from $\phi$ to $\psi$ in
$C_{T_1}$. Hence, $\phi\vdash\psi$ in $T_1$.

The next example shows that $F:T_1\to T_2$ being an equivalence does
not imply that $F$ is essentially surjective.

\begin{example} Let $T_1$ be the theory that says ``there are exactly
  two things,'' and let $T_2$ be the Morita extension of $T_1$ by the
  addition of a product sort. We know that the obvious embedding
  $F:T_1\to T_2$ is one-half of an equivalence. However, there are
  clearly formulas $\phi$ in the language of $T_2$ that are not
  provably equivalent to formulas in the target of the translation
  $F$. For example, let $\psi$ be the sentence that says there are
  four things of the product sort.

  Nonetheless, there is an extended sense in which this formula $\psi$
  is in the image of $F$. Indeed, if $\chi$ is the ``code'' for the
  product sort $\sigma '$, then for any $\Sigma _2$-formula $\psi (x)$
  with $x$ of type $\sigma '$, there is a $\Sigma _1$ formula
  $\phi (y)$ such that
  \[ T_2\: \vdash \: \chi (x,y)\to (\psi (x)\leftrightarrow \phi (x))
    .\] Notice that a code is really just another name for a t-map. I
  conjecture then that the correct formulation of ``essential
  surjectivity'' will necessarily involve t-maps. \end{example}

Let's now try to work at this in the other direction by looking at
abstract $2$-categoric notions of ``eso'' and ``fully faithful''
morphisms. See
\url{https://ncatlab.org/nlab/show/fully+faithful+morphism}

If $K$ is a 2-category, then for any objects $a,b$ of $K$, we let
$K(a,b)$ denote the category whose objects are $1$-cells (i.e. arrows)
from $a$ to $b$, and whose arrows are $2$-cells. Thus, in the
particular case of $\mathsf{Th}$, if $T_1$ and $T_2$ are theories,
then $\mathsf{Th}(T_1,T_2)$ is the category whose objects are
translations, and whose arrows are t-maps.

\begin{defn} In a $2$-category $\2K$, a $1$-cell $f:A\to B$ is said to
  be \emph{fully faithful} just in case for all objects $X\in \2K$,
  the induced functor $\2K (X,A)\to \2K (X,B)$ is full and
  faithful. \end{defn}

\begin{conj} Let $F:T_1\to T_2$ be a translation. Then $F$ is fully
  faithful iff $F$ is conservative. \end{conj}

NOTE: I now think this conjecture is false. I think that $F$ being
fully faithful is stronger than $F$ being conservative.

\bigskip TO DO: relate features of arrows in $\2{Th}$ to the Baez
classification of ``forgetting''
\citep{forget,weatherall2016,barrett2020}. Pay attention to the
ambiguity between syntactic and semantic arrows.

\section{Examples of theory pairs that are not CB}


\begin{enumerate}
\item $T_1$ is the empty theory on a countably infinite propositional
  signature. $T_2$ is the ``fan theory'' with axioms $p_0\vdash p_i$,
  for $i\geq 0$. These theories are counterexamples both to the CB and
  the co-CB properties.

  But are these theories ``pathological'' in some sense? The fact that
  these theories are propositional should not (I think) be seen as a
  pathology. However, these theories are incomplete, and the second of
  them is not finitely axiomatizable.
\item ZF and ZFC. See
  \url{https://cs.nyu.edu/pipermail/fom/2010-January/014325.html}
\item The many examples of pairs of set theories described in \citep[p
  8]{hamkins}.
\item The pair of theories described in \citep{andreka}.
\end{enumerate}

\section{Cantor-Bernstein}

The first thing I would like to show: for theories $T_1$ and $T_2$ to
violate Cantor-Bernstein, both have to have infinitely many
non-isomorphic models. (This fact is trivial in the case that $T_1$
and $T_2$ have models of infinite cardinality.)

\begin{note} Let $F:T_1\to T_2$ be a translation, and let
  $F^*:M(T_2)\to M(T_1)$ be the dual functor. Recall that:
  \begin{itemize}
  \item $F^*(M)\vDash \phi$ iff $M\vDash F(\phi )$, for any
    $\Sigma _1$-sentence $\phi$.
    \item $F^*$ is faithful by definition.
    \item $F^*$ reflects isomorphisms in the sense that if
      $F^*(j):F^*(M)\to F^*(N)$ is an isomorphism, then $j:M\to N$ is
      an isomorphism. This follows from the fact that an elementary
      embedding is an isomorphism iff it is surjective, and from the
      fact that the underlying function of $F^*(j)$ is none other than
      the underlying function of $j$.

      Note however: $F^*$ can still map non-isomorphic models to
      isomorphic models. For example: let $T_1$ be the theory that
      says that there are two things (in empty signature), let $T_2$
      be the theory that says that there are two things (in signature
      with a unary predicate symbol $P$), and let $F:T_1\to T_2$ be
      the inclusion. Let $M$ be a model in which the extension of $P$
      is empty, and let $N$ be a model in which the extension of $P$
      is non-empty. Then $F^*(M)$ is isomorphic to $F^*(N)$ although
      $M$ is not isomorphic to $N$.
    \end{itemize} \end{note}

\begin{defn} We say that a category $C$ is \emph{object-finite} just
  in case it has only finitely many objects up to
  isomorphism. \end{defn}

Let $T$ be a theory. If $M(T)$ is object-finite then the
L{\"o}wenheim-Skolem theorem implies that every model of $T$ has
finite cardinality. From this it follows that every hom set in $M(T)$
is finite. Furthermore, since for finite structures, elementary
equivalence implies isomorphism, it follows that for any
non-isomorphic $M,N$ in $M(T)$, there is a sentence $\phi$ such that
$M\vDash\phi$ and $N\vDash\neg\phi$.

\begin{prop} Let $T_1$ be a theory such that $M(T_1)$ is
  object-finite, and let $F:T_1\to T_2$ be a translation. If $F$ is
  conservative then $F^*:M(T_2)\to M(T_1)$ is essentially
  surjective. \end{prop}

\begin{proof} We prove the contrapositive. If $F^*:M(T_2)\to M(T_1)$
  is not eso, then there is a model $M$ of $T_1$ that is not
  isomorphic to any model of the form $F^*(N)$, with $N$ a model of
  $T_2$. By the preceding discussion, there is a sentence $\phi$ such
  that $M\vDash \phi$ but $F^*(N)\vDash\neg \phi$, for all models $N$
  of $T_2$. Hence $N\vDash F(\neg \phi )$ for all models $N$ of $T_2$,
  and by completeness, $T_2\vdash F(\neg \phi )$. Therefore $F$ is not
  conservative. \end{proof}

\begin{note} Let $C$ and $D$ be categories with respective object sets
  $C_0$ and $D_0$. Let $[C_0]$ and $[D_0]$ be the corresponding sets
  of equivalence classes of isomorphic objects. Each functor
  $F:C\to D$ induces a function $F_0:[C_0]\to [D_0]$, and $F_0$ is
  surjective iff $F$ is eso. If $F:C\to D$ and $G:D\to C$ are both
  eso, then Cantor-Bernstein for finite sets implies that $F_0$ is a
  bijection. \end{note}

\begin{lemma} Let $F$ be a finite set, let $f:F\to\mathbb{N}$ be a
  function, and let $\phi :F\to F$ be a bijection such that
  $f(x)\leq f(\phi (x))$ for all $x\in F$. Then $f(x)=f(\phi (x))$ for
  all $x\in F$. \end{lemma}

\begin{proof}[Sketch of proof] The function $f$ corresponds to a
  fibration of $F$ over $\mathbb{N}$. Since $\phi$ is a bijection, the
  size of the fibers remain constant, i.e.,
  $|f^{-1}(n)|=|(f\circ \phi )^{-1}(n)|$. Since $\phi$ is monotonic,
  it cannot move an element to a lower fiber. Thus no element can be
  moved out of the highest fiber, nor the next highest fiber,
  etc. \end{proof}

\begin{prop} Let $C$ and $D$ be totally-finite categories. If there
  are faithful, eso functors $F:C\to D$ and $G:D\to C$, then $C$ and
  $D$ are equivalent categories. In fact, $F$ itself is one half of an
  equivalence. \end{prop}

\begin{proof} By the above remark, $F_0:[C_0]\to [D_0]$ is a
  bijection. Since $F$ is automatically faithful, it will suffice to
  show that $F$ is full. For simplicity, we may henceforth replace $C$
  and $D$ with the corresponding skeletal categories.

  Consider the (finite) set $C_0\times C_0$ and the function
  $f:C_0\times C_0\to \mathbb{N}$ that assigns the cardinality of the
  corresponding hom set. That is, $f(a,b)=|\mathrm{hom}(a,b)|$. Let
  $g:D_0\times D_0\to\mathbb{N}$ be the corresponding function for
  $D$. Since $F$ is faithful, it induces a bijection
  $\eta :C_0\to D_0$ such that
  \[ f(a,b) \: \leq \: g(\eta (a),\eta (b)) .\] And since $G$ is
  faithful, it induces a bijection $\theta :D_0\to C_0$ such that
  \[ g(a,b) \: \leq \: f(\theta (a),\theta (b)) .\] If we let
  $\phi = \theta \circ\eta$ then
  \[ f(a,b) \: \leq \: f(\phi (a),\phi (b)) ,\] for all $a,b\in
  C_0$. By the above lemma, $f(a,b)=f(\phi (a),\phi (b))$, and it
  follows that $F$ is full. \end{proof}

\begin{conj} Suppose that $F:T_1\to T_2$ is a strong translation,
  i.e.\ one-dimensional, equality preserving, and with trivial domain
  formula. If $F$ is one-half of a weak equivalence, then $F$ is
  one-half of a strong equivalence. \end{conj}

\begin{proof}[Sketch of proof] We show that $F$ is conservative and
  essentially surjective. Let $G:T_2\to T_1$ be a translation such
  that $GF\simeq 1_{T_1}$ and $FG\simeq 1_{T_2}$. Let $\phi$ be a
  $\Sigma _1$-sentence such that $T_2\vdash F(\phi )$. Then
  $T_1\vdash GF(\phi )$, and (since the relevant t-map is trivial)
  $T_1\vdash \phi$. Therefore $F$ is conservative. Now let $\psi (x)$
  be a $\Sigma _2$-formula, for simplicity with one free
  variable. Then there a t-map $\chi (x,y)$ such that
  \[ T_2\: \vdash \: \chi (x,y)\to (\psi (x)\leftrightarrow (FG\psi
    )(y)) ,
  \] and $T_2$ implies that $\chi$ is a bijection on the domain. TO BE CONTINUED
 
\end{proof}

\begin{prop} Let $T_1$ and $T_2$ be proper theories such that $M(T_1)$
  and $M(T_2)$ are object-finite. Then $(T_1,T_2)$ has the
  Cantor-Bernstein property. \end{prop}

\begin{proof} Let $F:T_1\to T_2$ and $G:T_2\to T_1$ be conservative
  translations. By the previous results, $F^*:M(T_2)\to M(T_1)$ is
  part of an equivalence of categories. By Theorem 7.1 of
  \citep{summer2020}, $F$ is part of a homotopy
  equivalence.  \end{proof}

\begin{note} TO DO: I need to check the previous result. The problem
  is that the conclusion of Theorem 7.1 shows that $F$ is part of a
  weak equivalence. Does that automatically show that $F$ is part of a
  strong equivalence? Note that the dimension of $F$ is $1$, and the
  domain formula is trivial/universal. \end{note}

\begin{conj} If $M(T_1)$ and $M(T_2)$ are object-finite then
  $(T_1,T_2)$ has the Cantor-Bernstein property.  \end{conj}

\section{Co-Cantor-Bernstein}


\begin{prop} If $T_1$ or $T_2$ is complete, then $(T_1,T_2)$ has the
  co-CB property. \end{prop}

\begin{proof} If $T_1$ is complete, then every translation
  $F:T_1\to T_2$ is conservative. So if $F:T_1\to T_2$ is eso, then
  $F$ is a strong equivalence, i.e.\ $T_1$ and $T_2$ are
  bi-interpretable. \end{proof}

Recall that $F:T_1\to T_2$ is conservative iff $F^*:M(T_2)\to M(T_1)$
is a full functor \citep{barrett2020}. Recall also that if $F$ is a
strong (equality-preserving) translation, then $F^*$ preserves
cardinality of models. In particular, if $T_1$ and $T_2$ are
$\aleph _0$-categorical theories, then a translation $F:T_1\to T_2$
induces a group homomorphism
$F^*:\mathrm{Aut}(M_2)\to \mathrm{Aut}(M_1)$, and $F$ is conservative
iff $F^*$ is surjective. (In fact, $\mathrm{Aut}(M_i)$ is naturally a
topological group, and I conjecture that $F^*$ is a continuous group
homomorphism.)

Recall that if $T$ has countable signature, and if $T$ is
$\aleph _0$-categorical, then $T$ is complete. The following result
would be interesting because an $\aleph _0$-categorical theory is
essentially characterized by a topological group, viz.\ the group of
automorphisms of its unique (up to isomorphism) countable model.


\begin{conj} There are $\aleph _0$-categorical theories $T_1$ and
  $T_2$ that do not have the CB property. \end{conj}

\begin{defn} Let $I$ be the set of axioms
  $\{ \exists _{>1},\exists _{>2},\dots \}$. We say that a theory $T$
  is essentially finitely axiomatizable just in case there is a finite
  set $E$ of axioms such that $Cn(T)=Cn(E\cup I)$. \end{defn}

\begin{conj} There are essentially finitely axiomatizable theories
  $T_1$ and $T_2$ that do not have the CB property. \end{conj}

\section{Conjecture: co-CB fails only for theories with many models}

\begin{defn} For a theory $T$ and a cardinal number $\kappa$, let
  $I(T,\kappa )$ be the number of non-isomorphic countable models of
  $T$. \end{defn}

Interestingly, $I(T,\aleph _0)$ can be countably infinite, or any
finite cardinal besides $2$ \citep[see][p 155ff]{marker2006}.

\begin{prop} If $F:T_1\to T_2$ is essentially surjective, then for any
  fixed cardinal number $\kappa$,
  $I(T_2, \kappa )\leq I(T_1, \kappa )$. \end{prop}

\begin{proof} Recall that the dual functor $F^*$ is always
  faithful. If $F:T_1\to T_2$ is eso, then $F^*$ is also full
  \citep[Prop 6.6.13]{tlps}. In particular, for any models $M,N$ of
  $T_2$, if $F^*(M)$ is isomorphic to $F^*(N)$, then $M$ is isomorphic
  to $N$. Now fix a cardinal number $\kappa$, and let
  $[M(T_i)]_\kappa$ be the set of isomorphism classes of models of
  $T_i$ of cardinality $\kappa$. Then $F^*$ induces a one-to-one
  mapping from $[M(T_2)]_\kappa$ into $[M(T_1)]_\kappa$. \end{proof}

\begin{prop} Suppose that $T_2$ has finitely many non-isomorphic
  models of each cardinality. If $F:T_1\to T_2$ and $G:T_2\to T_1$ are
  essentially surjective, then $(F^*,G^*)$ is an equivalence of
  categories. If $T_1$ and $T_2$ are proper theories then $F^*$ is
  part of a homotopy equivalence. \end{prop}

\begin{proof} For the first part it will suffice to show that $F^*$ is
  essentially surjective. By the previous proof, $G^*$ induces an
  injection of $[M(T_1)]_\kappa$ into $[M(T_2)]_\kappa$. Since the
  latter is finite, so is the former. Since $F^*$ is an injection of
  one finite set into a not-larger finite set, it follows that $F^*$
  is bijection. Therefore $F^*$ is essentially surjective.

  The second part follows from Theorem 7.1 of
  \citep{summer2020}. \end{proof}

\begin{cor} Let $T_1$ and $T_2$ be proper theories. If $(T_1,T_2)$
  violate the co-CB property then there is a cardinal number $\kappa$
  such that $T_1$ and $T_2$ have infinitely many non-isomorphic models
  of size $\kappa$. \end{cor}

TO DO: I would like to come up with a similar necessary condition for
$T_1$ and $T_2$ violating the CB property. However, we do not yet have
any interesting result of the form: ``if $T_1$ or $T_2$ is \dots and
$F:T_1\to T_2$ is conservative then $F^*:M(T_2)\to M(T_1)$ is \dots
.''



\section{Groups that are not CB}

\begin{defn} Let $P(\mathbb{N})$ be the permutation group of the
  natural numbers, equipped with the topology of pointwise
  convergence. (TO DO: explain the sense in which this topology on
  $P(\mathbb{N})$ is definable from the theory of infinite
  sets. Explain more generally the sense in which for a
  $\Sigma$-structure $M$, $\mathrm{Aut}(M)$ is naturally a topological
  group.) \end{defn}

\begin{fact} Let $G$ be a subgroup of $P(\mathbb{N})$. Then $G$ is the
  automorphism group of an $\aleph _0$-categorical theory iff $G$ is
  a closed subset of $P(\mathbb{N})$. \end{fact}

\begin{conj} There are closed subgroups $G$ and $H$ of
  $P(\mathbb{N})$ such that $G$ is isomorphic to a closed subgroup of
  $H$ and vice versa, but $G$ and $H$ are not
  isomorphic. \end{conj}

It is not difficult at all to find groups that violate the
Cantor-Bernstein condition --- but I do not immediately know if any of
these groups are of the form $\mathrm{Aut}(M)$ for an
$\aleph _0$-categorical structure $M$.

\begin{enumerate}
\item The group $S_\infty$ of finite permutations of $\mathbb{N}$ and
  the alternating grooup $A_\infty$. See  \url{https://math.stackexchange.com/questions/1259081/if-there-are-injective-homomorphisms-between-two-groups-in-both-directions-are}
\item Infinite direct sums of $\mathbb{Z}_{2^i}$.
\item The free group on $2$ generators and the free group on $3$ generators.  
\end{enumerate}

\begin{prop}[\cite{ahlbrandt1986}] Two countable
  $\aleph _0$-categorical structures are bi-interpretable iff their
  automorphism groups are isomorphic as topological groups. \end{prop}

Conjecture: the previous result can be lifted to
$\aleph _0$-categorical theories (with countable signature). But we
need to be careful about terminology. First of all, Ahlbrandt and
Ziegler are working with a notion of ``interpretation'' between
structures of one language and structures of another language: given a
$\Sigma _1$-structure $\3M _1$ and a $\Sigma _2$-structure $\3M _2$,
and interpretation $f:\3M_1\to \3M_2$ consists of a surjection
$f:U\to M_2$ where $U$ is a definable subset of the domain of $\3M_1$
and $M_2$ is the domain of $\3M_2$, etc.

\begin{conj} An interpretation from $\3M_1$ to $\3M_2$ is a
  translation in our sense from $Th(\3M_1)$ to $Th(\3M_2)$. \end{conj}

The following proposition follows immediately from the fact that
$\aleph _0$-categorical theories are complete.

\begin{prop} If $T$ is $\aleph _0$-categorical with unique model
  $\3M$, then $T$ is logically equivalent to $Th(\3M )$. \end{prop}

This result would be interesting because it would show that the
structure of an $\aleph _0$-categorical theory is captured by its
countable model. i.e., there is no need to look at the models of
higher cardinality and the arrows between them.

\begin{prop}[\cite{evans1990}] There are closed subgroups $G$ and $H$
  of $P(\mathbb{N})$ that are isomorphic qua groups but not as
  topological groups. Hence, the corresponding structures are not
  bi-interpretable. \end{prop}

The former result is intriguing: there is a group isomorphism
$\varphi :G\xlongrightarrow{\sim} H$ that does not correspond to a
homotopy equivalence between the corresponding theories. In short: the
symmetries of the model are not enough to capture the structure of the
theories.

\section{$\aleph _0$-categorical theories}

For this discussion, we restrict to theories with countable
signatures. In this case, the downward L{\"o}wenheim-Skolem theorem
shows that an $\aleph _0$-categorical theory is complete.

Question: How can we characterize the syntactic categories of
$\aleph _0$-categorical theories? (Hint: Look at the Ryll-Nardzewski
theorem \url{https://en.wikipedia.org/wiki/Omega-categorical_theory}
and \citep[p 30]{cameron1990}. I suspect that the characterization
will have something to do with finiteness of subobject lattices.)

TO DO: Establish a correspondence between the $2$-category of
$\aleph _0$-categorical theories and some subcategory of the
$2$-category of topological groups. Note 1: every group is a category,
and group homomorphisms are functors.  Note 2: all translations
between such theories are conservative. It will be helpful to look at
pages 106--107 of \citep{cameron1990} where he describes conditions on
a topological group $G$ that ensure it corresponds to a categorical
theory.



\begin{defn} Given an $\aleph _0$-categorical theory $T$, let
  $\3G (T)$ be the topological group of automorphisms of its unique
  countable model. \end{defn}

\begin{conj} If $F:T_1\to T_2$ is a translation then
  $F^*|_{\3G (T_2)}$ is a continuous homomorphism from $\3G (T_2 )$ to
  $\3G (T_1)$. \end{conj}

\begin{conj} There is a 2-functor $\3G$ from the 2-category of
  topological groups to the 2-category of $\aleph _0$-categorical
  theories. (This is not stated correctly yet: we should not expect
  all topological groups to occur in the domain. It is only the
  ``nice'' topological groups, i.e.\ the automorphism groups of
  $\aleph _0$-categorical theories.) \end{conj}


\begin{conj} There is a 2-functor $\3F$ from the 2-category of
  $\aleph _0$-categorical theories and the 2-category of topological
  groups. \end{conj}







\printbibliography 



\end{document} 
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
